---
title: "PM566 Final Project"
author: "Hanke Zheng"
output:
  html_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---
**This is my PM566 Final Project website.**
```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(dtplyr)
library(ggplot2)
library(plotly)
library(DT)
library(knitr)

# Initialize code chunk options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
  class.source = "code-r")
```
# Introducation: 
**The objective is to explore (1)the association between disability prevalence and age, and (2)the association between prevalence of different types of disability across the US, especially between cognitive disability and mobidity disability. **
## Several specific questions were addressed.
- Is the prevalence of distributed evenly across states?
- On average, what is the mean prevalence for each disability in each age group? 
- Within each age group, which state has the highest/lowest overall prevalence of any disability? How about cognitive disability and mobidity disability? 
- Within each age group, what type of disability is the most prevalent across all the states, and which one is the leaste prevalent? 
- Is there an association between age and prevalence of disability? 
- Is there an association between prevalence of cognitive disability and mobidity disability? 

## Background of the raw dataset
The data source for this program is the Disability and Health Data System (DHDS), which is an online state-level data on adults with disabilities in the US in 2018. Prevalence on six functional disability types, including cognitive (serious difficulty concentrating, remembering or making decisions), hearing (serious difficulty hearing or deaf), mobility (serious difficulty walking or climbing stairs), vision (serious difficulty seeing), self-care (difficulty dressing or bathing) and independent living (difficulty doing errands alone),are reported by age group, race, gender, and veteran status within each state, respectively. 

## Methods:
### Data Source
- The data was downloaded from the Center of Disease Control and Prevention(https://data.cdc.gov/).
- There are 7168 rows and 31 coloumns in the raw dataset; each raw gives a piece of information of each each state. It is not ordered. 
- The raw dataset is a long dataset, with the prevalence of each type of disbility and the prevalence of any disbility by reponse type (e.g. age, race, gender, veteran status) for each state and Health and 11 Humen Health Service (HHS) regions reported in each row. The prevalence of the 6 disability types doesn't add up equal to the prevalence of any disability, the potential explanation is that a proportion of people might have multiple conditions.
```{r, message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
library(data.table)
### First download the data and read into data table
dhds <- data.table::fread("/Users/hankezheng/pm512/PM-566-my-final-project/data/DHDS.csv")
dhds <- data.frame(dhds)
dhds_dt <- data.table(dhds)
dim(dhds)
head(dhds)
tail(dhds)
str(dhds)
### Data cleaning & check on important variables
library(dplyr)
library(dtplyr)
table(dhds_dt$LocationAbbr)
table(dhds_dt$Response)
table(dhds_dt$Stratification1)
table(dhds_dt$StratificationCategory1)
summary(dhds_dt$Data_Value)
#Select important variables and subset to only include data points reported for each state by age group 
dhds_age <- dhds %>% filter(IndicatorID=="AGEIND") 
dhds_age <- dhds_age %>% select(LocationAbbr, LocationDesc, Response, Data_Value,
                      Low_Confidence_Limit, High_Confidence_Limit, Number, 
                      StratificationCategory1, Stratification1, IndicatorID, Geolocation)
# Filter out rows of HHS regions
dhds %>% 
  filter(!grepl('HHS', LocationAbbr)) %>% 
  compute() %>% 
  nrow()
dhds_age <- dhds_age %>% filter(!grepl('HHS', LocationAbbr))
# Check missing values for the key variable 
dhds_age_dt <- data.table(dhds_age)
dim(dhds_age_dt[is.na(Data_Value)])[1]/dim(dhds_age_dt)[1]
# Rename them to make things easier for further analysis
names(dhds_age)[names(dhds_age) == "Response"] <- "age_grp"
names(dhds_age)[names(dhds_age) == "Stratification1"] <- "dis_type"
names(dhds_age)[names(dhds_age) == "LocationAbbr"] <- "state"
names(dhds_age)[names(dhds_age) == "Data_Value"] <- "prevalence"
table(dhds_age$age_grp)
table(dhds_age$dis_type)
table(dhds_age$state)
summary(dhds_age$prevalence)
```

### Data Preparation
- R package "data table", "dplyr", and "dplyer" were mainly used to check, and manipulate the data to create a final main dataset for further analysis. 
- To clean the dataset to answer the research question, I excluded irrelevant rows for which the 'responses' were not age and the values reported from the HHS level. Some variables were renamed for easier reference.
- For future comparisons between prevalence of different types of disabilities, the dataset was reshaped from long to wide to be a state-level dataset, with the prevalence of different disabilities listed as seperate colounms. 
- Comparing the prevalence value in each state as a raw, a new categorical variable was created to record the disease type with the greatest prevalence for each state by age. 
- In the final main dataset for analysis, there are 162 rows/observations (each row gives the statistics of each state or region divided by different age groups) and 13 coloumns/variables of interest. 
- The number of NAs for those key variables (prevalence) is very small. 
```{r, message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
# Reshape the data from long to wide 
# Making sure the main dataset is sorted by state and age group. 
dhds_age_any <- dhds_age %>% filter(grepl('Any', dis_type)) %>% arrange(state, age_grp) 
names(dhds_age_any)[names(dhds_age_any) == "prevalence"] <- "pre_any"
summary(dhds_age_any$pre_any)

dhds_age_cog <- dhds_age %>% filter(grepl('Cog', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_cog)[names(dhds_age_cog) == "prevalence"] <- "pre_cog"
dhds_age_cog <- dhds_age_cog %>% select(pre_cog)

dhds_age_hearing <- dhds_age %>% filter(grepl('Hear', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_hearing)[names(dhds_age_hearing) == "prevalence"] <- "pre_hear"
dhds_age_hearing <- dhds_age_hearing %>% select(pre_hear)

dhds_age_mob <- dhds_age %>% filter(grepl('Mob', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_mob)[names(dhds_age_mob) == "prevalence"] <- "pre_mob"
dhds_age_mob <- dhds_age_mob %>% select(pre_mob)

dhds_age_vis <- dhds_age %>% filter(grepl('Vis', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_vis)[names(dhds_age_vis) == "prevalence"] <- "pre_vis"
dhds_age_vis <- dhds_age_vis %>% select(pre_vis)

dhds_age_self <- dhds_age %>% filter(grepl('Self', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_self)[names(dhds_age_self) == "prevalence"] <- "pre_self"
dhds_age_self <- dhds_age_self %>% select(pre_self)

dhds_age_ind <- dhds_age %>% filter(grepl('Ind', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_ind)[names(dhds_age_ind) == "prevalence"] <- "pre_ind"
dhds_age_ind <- dhds_age_ind %>% select(pre_ind)

dhds_age_comb <- cbind(dhds_age_any,dhds_age_cog, dhds_age_hearing, dhds_age_mob, dhds_age_vis, dhds_age_self, dhds_age_ind)

# Created a new variable to document the disease type with the greatest prevalence for each state by age.
max_pre <- dhds_age_comb %>% select(pre_cog,pre_hear,pre_mob,pre_vis,pre_self,pre_ind)
max_pre$max<-apply(X=max_pre, MARGIN=1, FUN=max, na.rm=TRUE )
head(max_pre)
max_pre <- max_pre %>% select(max)
dhds_age_comb <- cbind(dhds_age_comb,max_pre)
head(dhds_age_comb)
dhds_age_comb$most_dis_type <- ifelse(dhds_age_comb$max==dhds_age_comb$pre_cog, "cognitive", 
                               ifelse(dhds_age_comb$max==dhds_age_comb$pre_hear, "hearing",
                                ifelse(dhds_age_comb$max==dhds_age_comb$pre_mob, "mobidity", 
                                ifelse(dhds_age_comb$max==dhds_age_comb$pre_vis,"vision",
                                ifelse(dhds_age_comb$max==dhds_age_comb$pre_self, "self care","independence")))))
table(dhds_age_comb$most_dis_type)
# Only keep important variables in the final dataset for analysis
str(dhds_age_comb)
dhds_age_comb <- dhds_age_comb %>% select(state, LocationDesc,age_grp, pre_any,
                                   pre_cog, pre_hear, pre_mob, pre_vis,pre_self, pre_ind,
                                   most_dis_type)
str(dhds_age_comb)
head(dhds_age_comb)
```

### Exploratory Data Analysis
```{r, message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
# Check if there's any NA for each prevalence value and do some exploratory statistics 
summary(dhds_age_comb$pre_any)
summary(dhds_age_comb$pre_cog)
summary(dhds_age_comb$pre_hear)
summary(dhds_age_comb$pre_mob)
summary(dhds_age_comb$pre_vis)
summary(dhds_age_comb$pre_self)
summary(dhds_age_comb$pre_ind)
hist(dhds_age_comb$pre_any)
hist(dhds_age_comb$pre_cog)
hist(dhds_age_comb$pre_hear)
hist(dhds_age_comb$pre_mob)
hist(dhds_age_comb$pre_vis)
hist(dhds_age_comb$pre_self)
hist(dhds_age_comb$pre_ind)
# Which state has the greatest prevalence of any disability and which state has the least prevalence?
library(data.table)
dhds_age_comb_dt <- data.table(dhds_age_comb)
dhds_age_comb_dt[(age_grp=="18-44")][order(pre_any)]%>% head(1)
dhds_age_comb_dt[(age_grp=="18-44")][order(pre_any)]%>% tail(1)
dhds_age_comb_dt[(age_grp=="45-64")][order(pre_any)]%>% head(1)
dhds_age_comb_dt[(age_grp=="45-64")][order(pre_any)]%>% tail(1)
dhds_age_comb_dt[(age_grp=="65+")][order(pre_any)]%>% head(1)
dhds_age_comb_dt[(age_grp=="65+")][order(pre_any)]%>% tail(1)

```
- The number of NAs for those key variables (prevalence) is very small. 
- The distribution of prevalence of any disease is not normally distributed. With a mean of 30.92%, most data points are concentrated between 15-25% and 40-45%. Except for hearing and mobilidity disabilities, the distribution of all other types of disability is normal. The mean of prevalence of mobility disability is the highest (16.6%) whereas it's the lowest for self-care disability (4.365%).
- For age 18-44, disability of any disease is the least prevalent in DC (12.9%) whereas it's most prevalent in Puerto Rico(29.3%). For age 45-65, For age 65+,disability of any disease is the least prevalent in Colorado (20.6%) whereas it's the most prevalent in Puerto Rico(53.3%). For 65+, disability of any disease is the least prevalent in Colorado (32.2%) whereas it's the most prevalent in Puerto Rico (62.8%).

## Results
**Table of Average Prevalence of Disability by Age**
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(data.table)
dhds_age_comb_dt <- data.table(dhds_age_comb_dt)
dhds_age_avg <- dhds_age_comb_dt[, .(
    pre_any_avg = mean(pre_any, na.rm=TRUE),
    pre_cog_avg = mean(pre_cog, na.rm=TRUE),
    pre_hear_avg=mean(pre_hear, na.rm=TRUE), 
    pre_mob_avg=mean(pre_mob, na.rm=TRUE), 
    pre_vis_avg=mean(pre_vis,na.rm=TRUE),
    pre_self_avg=mean(pre_self,na.rm=TRUE), 
    pre_ind_avg=mean(pre_ind,na.rm=TRUE)
    ),
    by = c("age_grp")
    ][order(age_grp)] %>% head()
names(dhds_age_avg)[names(dhds_age_avg) == "age_grp"] <- "Age"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_any_avg"] <- "Any disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_cog_avg"] <- "Cognitive disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_hear_avg"] <- "Hearing disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_mob_avg"] <- "Mobidity disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_vis_avg"] <- "Vision disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_self_avg"] <- "Self care disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_ind_avg"] <- "Independent disability"
# knit a table for the results
library(knitr)
kable(dhds_age_avg, format="simple")
```

```{r, message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
library(ggplot2)
library(plotly)
dt1 <- dhds_age_comb %>%
  filter(!(pre_any %in% NA))
p1 <-ggplot(dt1, 
           aes(y=pre_any, fill=age_grp)
           ) + geom_boxplot() +
           facet_wrap(~age_grp,nrow=1) +
           labs(title="Prevalence of any disability by age", y="prevalence of any disability(%)")
box_1 <- ggplotly(p1)

# Histogram of prevalence of any disease 
p4<-ggplot(dhds_age_comb)+
  geom_histogram(mapping = aes(x=pre_any, fill=age_grp))+
  scale_fill_discrete(name="Age")+
  scale_fill_brewer(palette = "Spectral")+
  labs(title="Distribution of Prevalence of Any Disability", 
       x="Prevalence(%)", y="Count") +
  theme()
hist_1 <- ggplotly(p4)

# Histogram of prevalence of cognitive disease 
p6<-ggplot(dhds_age_comb)+
  geom_histogram(mapping = aes(x=pre_cog, fill=age_grp))+
  scale_fill_discrete(name="Age")+
  scale_fill_brewer(palette = "Spectral")+
  labs(title="Distribution of Prevalence of Cognitive Disability", 
       x="Prevalence(%)", y="Count") +
  theme()
hist_2 <- ggplotly(p6)
# Histogram of prevalence of mobidity disease 
p7<-ggplot(dhds_age_comb)+
  geom_histogram(mapping = aes(x=pre_mob, fill=age_grp))+
  scale_fill_discrete(name="Age")+
  scale_fill_brewer(palette = "Spectral")+
  labs(title="Distribution of Prevalence of Mobidity Disability", 
       x="Prevalence(%)", y="Count") +
  theme()
hist_3 <- ggplotly(p7)

# Histogram of the most prevalent disease by age group
p5 <- ggplot(dhds_age_comb)+
  geom_bar(mapping=aes(x=age_grp,fill=most_dis_type))+
  # change the default palette
  scale_fill_brewer(palette = "Spectral")+
  labs(title="Most prevalent disability type by age group", x="age group", y="Count")+
  theme()
hist2 <- ggplotly(p5)

# Explore the association between prevalence of different types of disabilities. 
dt3 <- dhds_age_comb %>%filter(!(pre_mob%in% NA) & !(pre_cog%in% NA))
p3<- ggplot(dt3, mapping=aes(x=pre_cog, y=pre_mob, color=age_grp))+
  geom_point()+
  stat_smooth(method=lm) +
  labs(title="The association between prevalence of cognitive and mobidity disability", x="cognitive disability(%)", y="mobidity disability(%)")
  theme()
scat_1 <- ggplotly(p3)

# Create maps for each age group to see if prevalence of disease is evenly distrivuted across states. 
## Create hover text
dhds_age_comb$hover <- with(dhds_age_comb, paste(LocationDesc, '<br>', "Prevalence of cognitive disability per 100: ", pre_cog, '<br>', "Prevalence of mobidity disability per 100: ", pre_mob, '<br>', "Prevalence of any disability per 100: ", pre_any))

dt_18 <- dhds_age_comb %>% filter(dhds_age_comb$age_grp=="18-44")
dt_45 <- dhds_age_comb %>% filter(dhds_age_comb$age_grp=="45-64")
dt_65 <- dhds_age_comb %>% filter(dhds_age_comb$age_grp=="65+")

## Setting up map to present the prevalence of all disability
set_map <- list(
  scope = 'usa',
  projection = list(type = 'conic')
)
## Create the map
map_18 <- plot_geo(dt_18, locationmode = 'USA-states') %>% 
        add_trace(
          z = ~pre_any, text=~hover, locations = ~state,
          color=~pre_any, colors='Greens'
          ) %>%
        colorbar(title = "Prevalence of any disability per 100")%>%
        layout(title="Prevalence of Disability by State for Age < 18",
               geo=set_map)

map_45 <- plot_geo(dt_45, locationmode = 'USA-states') %>% 
        add_trace(
          z = ~pre_any, text=~hover, locations = ~state,
          color=~pre_any, colors='Greens'
          ) %>%
        colorbar(title = "Prevalence of any disability per 100")%>%
        layout(title="Prevalence of Disability by State for Age 18-45",
               geo=set_map)

map_65 <- plot_geo(dt_65, locationmode = 'USA-states') %>% 
        add_trace(
          z = ~pre_any, text=~hover, locations = ~state,
          color=~pre_any, colors='Greens'
          ) %>%
        colorbar(title = "Prevalence of any disability per 100")%>%
        layout(title="Prevalence of Disability by State for Age > 65",
               geo=set_map)
```


```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p1) 
```
- For age of 18-44, the mean (range) of the prevalence of any disability across the nation is 18.7% (12.9-29.3%).
- For age of 45-65, the mean (range) of the prevalence of any disability across the nation is 28.1% (20.6-53.3%).
- For age over 65, the mean (range) of the prevalence of any disability across the nation is 40.3% (32.2-62.8%).

### Distribution of Prevalence of Disability{.tabset}
#### Any Disability
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p4)
```

#### Cognitive Disability
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p6)
```

#### Mobidity Disability
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p7)
```
**Will elaborate more here for the narrative**
- For the prevalence of any disability, it is normally distributed for age 18-44 and age over 65 whereas it's not normally distributed for age of 44-65. 
- The prevalence of any disability is concentrated on the left for age 18-44, with the most prevalent value being 20.6% (N=10 states).
- The prevalence of any disability is concentrated in the middle for age 45-64, with the most prevalent value being 39.6% (N=10 states).
- The prevalence of any disability is concentrated on the right for age over 65, with the most prevalent value being 43% (N=10 states).

```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p5) 
```
The most prevalent disability across the nation is cognitive disability for the young population (aged 18-44) in all states, and mobidity disability is the most prevalent for older population. 

```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p3)
```
A positive association between prevalence of cognitive disability and mobidity disability is observed in all age groups. The slope is the flattest in the young population and it gets sharper in older population. 

### Distribution of Disability by State {.tabset}
#### For age of 18-44
```{r, message=FALSE, echo=FALSE, warning=FALSE}
map_18
```

#### For age of 45-65
```{r, message=FALSE, echo=FALSE, warning=FALSE}
map_45
```


#### For age over 65
```{r, message=FALSE, echo=FALSE, warning=FALSE}
map_65
```

- Based on the calculated average prevalence by age, there is an increasing trend by age seen in the prevalence of any disability: 18.86% for 18-44, 29.77% for 45-64, and 44.13% for 65+. Similar increasing trends are observed in hearing disability, mobility disability, vision disability and independence disability. An obvious positive association between age and prevalence of any disability was osberved from the boxplots. 

- As shown from the barchart, the prevalence of cognitive disability is the greatest for age of 18-44 in all states in 2018. Mobidity disability is the most prevalent for age 45-64 in all states, and it's also the most prevalent for age 65+ in most states (the prevalence of hearing disability is the greatest in several states).

- As shown from the scatterplots, there is a positive association between prevalence of cognitive disability and mobidity disability, meaning that if the prelence of cognitive disability is high in one state the prevalence of mobidity disability is also likely to be high.The slopes for age 45-64 and 65+ are close whereas the curve is relatively flat for age of 18-44. 

## Conclusion:
- Puerto Rico is the region with the highest prevalence of any disability in all ages in 2018.
- Across all states, cognitive disability and mobidity disability are the two that are most prevalent in comparative to other types of disability, with cognitive disability more prevalent among the young (<65) and mobidity disability being more common in older population(65+). 
- Overall, the prevalences of any disability, hearing disability, mobility disability, vision disability and independence disability are positively associated with aging. People are more likely to be in a stage with these types of disability as they age. 
- There is a positive association between prevalence of cognitive disability and mobidity disability, but mroe information is needed to ascertain this relationship. 


