---
title: "Explore the Prevalence of US Adult Disability in 2018"
author: "Hanke Zheng"
output:
  html_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
  pdf_document:
    toc: yes
---
**Email address: hankezhe@usc.edu**

**Written report: https://github.com/hankezhe/PM-566-my-final-project/blob/main/written%20report.pdf **

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(dtplyr)
library(ggplot2)
library(plotly)
library(DT)
library(knitr)

# Initialize code chunk options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px",
  class.source = "code-r")
```
**This website is to take an overview of the prevalence of disability by region and age among US adults using 2018 data derived from the Disability and Health Data System (DHDS). All datapoints for this analysis are actual states/regions in the US. **

```{r, message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
library(data.table)
### First download the data and read into data table
dhds <- data.table::fread("/Users/hankezheng/pm512/PM-566-my-final-project/data/DHDS.csv")
dhds <- data.frame(dhds)
dhds_dt <- data.table(dhds)
dim(dhds)
head(dhds)
tail(dhds)
str(dhds)
### Data cleaning & check on important variables
library(dplyr)
library(dtplyr)
table(dhds_dt$LocationAbbr)
table(dhds_dt$Response)
table(dhds_dt$Stratification1)
table(dhds_dt$StratificationCategory1)
summary(dhds_dt$Data_Value)
#Select important variables and subset to only include data points reported for each state by age group 
dhds_age <- dhds %>% filter(IndicatorID=="AGEIND") 
dhds_age <- dhds_age %>% select(LocationAbbr, LocationDesc, Response, Data_Value,
                      Low_Confidence_Limit, High_Confidence_Limit, Number, 
                      StratificationCategory1, Stratification1, IndicatorID, Geolocation)
# Filter out rows of HHS regions
dhds %>% 
  filter(!grepl('HHS', LocationAbbr)) %>% 
  compute() %>% 
  nrow()
dhds_age <- dhds_age %>% filter(!grepl('HHS', LocationAbbr))
# Check missing values for the key variable 
dhds_age_dt <- data.table(dhds_age)
dim(dhds_age_dt[is.na(Data_Value)])[1]/dim(dhds_age_dt)[1]
# Rename them to make things easier for further analysis
names(dhds_age)[names(dhds_age) == "Response"] <- "age_grp"
names(dhds_age)[names(dhds_age) == "Stratification1"] <- "dis_type"
names(dhds_age)[names(dhds_age) == "LocationAbbr"] <- "state"
names(dhds_age)[names(dhds_age) == "Data_Value"] <- "prevalence"
table(dhds_age$age_grp)
table(dhds_age$dis_type)
table(dhds_age$state)
summary(dhds_age$prevalence)
```

```{r, message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
# Reshape the data from long to wide 
# Making sure the main dataset is sorted by state and age group. 
dhds_age_any <- dhds_age %>% filter(grepl('Any', dis_type)) %>% arrange(state, age_grp) 
names(dhds_age_any)[names(dhds_age_any) == "prevalence"] <- "pre_any"
summary(dhds_age_any$pre_any)

dhds_age_cog <- dhds_age %>% filter(grepl('Cog', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_cog)[names(dhds_age_cog) == "prevalence"] <- "pre_cog"
dhds_age_cog <- dhds_age_cog %>% select(pre_cog)

dhds_age_hearing <- dhds_age %>% filter(grepl('Hear', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_hearing)[names(dhds_age_hearing) == "prevalence"] <- "pre_hear"
dhds_age_hearing <- dhds_age_hearing %>% select(pre_hear)

dhds_age_mob <- dhds_age %>% filter(grepl('Mob', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_mob)[names(dhds_age_mob) == "prevalence"] <- "pre_mob"
dhds_age_mob <- dhds_age_mob %>% select(pre_mob)

dhds_age_vis <- dhds_age %>% filter(grepl('Vis', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_vis)[names(dhds_age_vis) == "prevalence"] <- "pre_vis"
dhds_age_vis <- dhds_age_vis %>% select(pre_vis)

dhds_age_self <- dhds_age %>% filter(grepl('Self', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_self)[names(dhds_age_self) == "prevalence"] <- "pre_self"
dhds_age_self <- dhds_age_self %>% select(pre_self)

dhds_age_ind <- dhds_age %>% filter(grepl('Ind', dis_type)) %>% arrange(state, age_grp)
names(dhds_age_ind)[names(dhds_age_ind) == "prevalence"] <- "pre_ind"
dhds_age_ind <- dhds_age_ind %>% select(pre_ind)

dhds_age_comb <- cbind(dhds_age_any,dhds_age_cog, dhds_age_hearing, dhds_age_mob, dhds_age_vis, dhds_age_self, dhds_age_ind)

# Created a new variable to document the disease type with the greatest prevalence for each state by age.
max_pre <- dhds_age_comb %>% select(pre_cog,pre_hear,pre_mob,pre_vis,pre_self,pre_ind)
max_pre$max<-apply(X=max_pre, MARGIN=1, FUN=max, na.rm=TRUE )
head(max_pre)
max_pre <- max_pre %>% select(max)
dhds_age_comb <- cbind(dhds_age_comb,max_pre)
head(dhds_age_comb)
dhds_age_comb$most_dis_type <- ifelse(dhds_age_comb$max==dhds_age_comb$pre_cog, "cognitive", 
                               ifelse(dhds_age_comb$max==dhds_age_comb$pre_hear, "hearing",
                                ifelse(dhds_age_comb$max==dhds_age_comb$pre_mob, "mobidity", 
                                ifelse(dhds_age_comb$max==dhds_age_comb$pre_vis,"vision",
                                ifelse(dhds_age_comb$max==dhds_age_comb$pre_self, "self care","independence")))))
table(dhds_age_comb$most_dis_type)
# Only keep important variables in the final dataset for analysis
str(dhds_age_comb)
dhds_age_comb <- dhds_age_comb %>% select(state, LocationDesc,age_grp, pre_any,
                                   pre_cog, pre_hear, pre_mob, pre_vis,pre_self, pre_ind,
                                   most_dis_type)
str(dhds_age_comb)
head(dhds_age_comb)
```


```{r, message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
# Check if there's any NA for each prevalence value and do some exploratory statistics 
summary(dhds_age_comb$pre_any)
summary(dhds_age_comb$pre_cog)
summary(dhds_age_comb$pre_hear)
summary(dhds_age_comb$pre_mob)
summary(dhds_age_comb$pre_vis)
summary(dhds_age_comb$pre_self)
summary(dhds_age_comb$pre_ind)
hist(dhds_age_comb$pre_any)
hist(dhds_age_comb$pre_cog)
hist(dhds_age_comb$pre_hear)
hist(dhds_age_comb$pre_mob)
hist(dhds_age_comb$pre_vis)
hist(dhds_age_comb$pre_self)
hist(dhds_age_comb$pre_ind)
# Which state has the greatest prevalence of any disability and which state has the least prevalence?
library(data.table)
dhds_age_comb_dt <- data.table(dhds_age_comb)
dhds_age_comb_dt[(age_grp=="18-44")][order(pre_any)]%>% head(1)
dhds_age_comb_dt[(age_grp=="18-44")][order(pre_any)]%>% tail(1)
dhds_age_comb_dt[(age_grp=="45-64")][order(pre_any)]%>% head(1)
dhds_age_comb_dt[(age_grp=="45-64")][order(pre_any)]%>% tail(1)
dhds_age_comb_dt[(age_grp=="65+")][order(pre_any)]%>% head(1)
dhds_age_comb_dt[(age_grp=="65+")][order(pre_any)]%>% tail(1)

```

### Table of Average Prevalence of Disability by Age
```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(data.table)
dhds_age_comb_dt <- data.table(dhds_age_comb_dt)
dhds_age_avg <- dhds_age_comb_dt[, .(
    pre_any_avg = mean(pre_any, na.rm=TRUE),
    pre_cog_avg = mean(pre_cog, na.rm=TRUE),
    pre_hear_avg=mean(pre_hear, na.rm=TRUE), 
    pre_mob_avg=mean(pre_mob, na.rm=TRUE), 
    pre_vis_avg=mean(pre_vis,na.rm=TRUE),
    pre_self_avg=mean(pre_self,na.rm=TRUE), 
    pre_ind_avg=mean(pre_ind,na.rm=TRUE)
    ),
    by = c("age_grp")
    ][order(age_grp)] %>% head()
names(dhds_age_avg)[names(dhds_age_avg) == "age_grp"] <- "Age"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_any_avg"] <- "Any disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_cog_avg"] <- "Cognitive disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_hear_avg"] <- "Hearing disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_mob_avg"] <- "Mobidity disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_vis_avg"] <- "Vision disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_self_avg"] <- "Self care disability"
names(dhds_age_avg)[names(dhds_age_avg) == "pre_ind_avg"] <- "Independent disability"
# knit a table for the results
library(knitr)
kable(dhds_age_avg, format="simple")
```

```{r, message=FALSE, echo=FALSE, warning=FALSE, results='hide',fig.show='hide'}
library(ggplot2)
library(plotly)
dt1 <- dhds_age_comb %>%
  filter(!(pre_any %in% NA))
p1 <-ggplot(dt1, 
           aes(y=pre_any, fill=age_grp)
           ) + geom_boxplot() +
           facet_wrap(~age_grp,nrow=1) +
           labs(y="Prevalence(%)",fill="Age")
box_1 <- ggplotly(p1)

# Histogram of prevalence of any disease 
p4<-ggplot(dhds_age_comb)+
  geom_histogram(mapping = aes(x=pre_any, fill=age_grp))+
  scale_fill_brewer(palette = "Spectral")+
  labs(title="Distribution of Prevalence of Any Disability", 
       x="Prevalence(%)", y="Count",fill="Age") +
  theme()
ggplotly(p4)

# Histogram of prevalence of cognitive disease 
p6<-ggplot(dhds_age_comb)+
  geom_histogram(mapping = aes(x=pre_cog, fill=age_grp))+
  scale_fill_brewer(palette = "Spectral")+
  labs(title="Distribution of Prevalence of Cognitive Disability", 
       x="Prevalence(%)", y="Count", fill="Age") +
  theme()
hist_2 <- ggplotly(p6)
# Histogram of prevalence of mobidity disease
p7<-ggplot(dhds_age_comb)+
  geom_histogram(mapping = aes(x=pre_mob, fill=age_grp))+
  scale_fill_brewer(palette = "Spectral")+
  labs(title="Distribution of Prevalence of Mobidity Disability", 
       x="Prevalence(%)", y="Count",fill="Age") +
  theme()
hist_3 <- ggplotly(p7)

# Histogram of the most prevalent disease by age group
p5 <- ggplot(dhds_age_comb)+
  geom_bar(mapping=aes(x=age_grp,fill=most_dis_type))+
  # change the default palette
  scale_fill_brewer(palette = "Spectral")+
  labs(title="Most prevalent disability type by age group", x="age group", y="Count",
       fill="Most Prevalent Disability")+
  theme()
hist2 <- ggplotly(p5)


# Explore the association between prevalence of different types of disabilities. 
dt3 <- dhds_age_comb %>%filter(!(pre_mob%in% NA) & !(pre_cog%in% NA))
p3<- ggplot(dt3, mapping=aes(x=pre_cog, y=pre_mob, color=age_grp))+
  geom_point()+
  stat_smooth(method=lm) +
  scale_colour_discrete("Age")+
  labs(x="cognitive disability(%)", y="mobidity disability(%)")+
  theme()
scat_1 <- ggplotly(p3)

# Create maps for each age group to see if prevalence of disease is evenly distrivuted across states. 
## Create hover text
dhds_age_comb$hover <- with(dhds_age_comb, paste(LocationDesc, '<br>', "Prevalence of cognitive disability per 100: ", pre_cog, '<br>', "Prevalence of mobidity disability per 100: ", pre_mob, '<br>', "Prevalence of any disability per 100: ", pre_any))

dt_18 <- dhds_age_comb %>% filter(dhds_age_comb$age_grp=="18-44")
dt_45 <- dhds_age_comb %>% filter(dhds_age_comb$age_grp=="45-64")
dt_65 <- dhds_age_comb %>% filter(dhds_age_comb$age_grp=="65+")

## Setting up map to present the prevalence of all disability
set_map <- list(
  scope = 'usa',
  projection = list(type = 'conic')
)
## Create the map
map_18 <- plot_geo(dt_18, locationmode = 'USA-states') %>% 
        add_trace(
          z = ~pre_any, text=~hover, locations = ~state,
          color=~pre_any, colors='Greens'
          ) %>%
        colorbar(title = "Prevalence of any disability per 100")%>%
        layout(title="Prevalence of Disability by State for Age < 18",
               geo=set_map)

map_45 <- plot_geo(dt_45, locationmode = 'USA-states') %>% 
        add_trace(
          z = ~pre_any, text=~hover, locations = ~state,
          color=~pre_any, colors='Greens'
          ) %>%
        colorbar(title = "Prevalence of any disability per 100")%>%
        layout(title="Prevalence of Disability by State for Age 18-45",
               geo=set_map)

map_65 <- plot_geo(dt_65, locationmode = 'USA-states') %>% 
        add_trace(
          z = ~pre_any, text=~hover, locations = ~state,
          color=~pre_any, colors='Greens'
          ) %>%
        colorbar(title = "Prevalence of any disability per 100")%>%
        layout(title="Prevalence of Disability by State for Age > 65",
               geo=set_map)
```

### Boxplot Comparing Prevalence of Any Disability by Age
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p1) 
```

- For age of 18-44, the mean (range) of the prevalence of any disability across the nation is 18.7% (12.9-29.3%).
- For age of 45-65, the mean (range) of the prevalence of any disability across the nation is 28.1% (20.6-53.3%).
- For age over 65, the mean (range) of the prevalence of any disability across the nation is 40.3% (32.2-62.8%).

### Distribution of Prevalence of Disability{.tabset}
*The data could be spread out due to the small sample size.*

- For the prevalence of any disability, it is normally distributed for age 18-44 and age over 65 whereas it's not normally distributed for age of 44-65. 
- The prevalence of any disability is concentrated on the left for age 18-44, with the most prevalent value being 20.6% (N=10 states).
- The prevalence of any disability is concentrated in the middle for age 45-64, with the most prevalent value being 39.6% (N=10 states).
- The prevalence of any disability is concentrated on the right for age over 65, with the most prevalent value being 43% (N=10 states).
- For age 18-44, the prevalence of cognitive disability is all under 10%, most states/regions are under 5% (N=25).
- For age 45-64,the most prevalent value being around 12.5% and around 16% (N=9 states/regions for each).
- Fpr age > 65, the prevalence of cognitive disability is concentrated between 22% to 28%, with the most prevalent value being around 27% (N=9 states/resgions).
- The distribution of prevalence of mobidity disability is more concentrated in comparative to that of cognitive disability. 
- For age 18-44, the prevalence of cognitive disability is all under 10%, most states/regions are under 5% (N=25).
- For age 45-64,the most prevalent value being around 12.5% and around 16% (N=9 states/regions for each).
- The prevalence of cognitive disability is concentrated between 22% to 28%, with the most prevalent value being around 27% (N=9 states/resgions).

#### Any Disability
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p4)
```

#### Cognitive Disability
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p6)
```

#### Mobidity Disability
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p7)
```

### What is the most prevalent disability across the nation? 
The most prevalent disability across the nation is cognitive disability for the young population (aged 18-44) in all states, and mobidity disability is the most prevalent for older population. 
```{r, message=FALSE, echo=FALSE, warning=FALSE}
ggplotly(p5) 
```


### Association between prevalence of cognitive and mobidity disability
A positive association between prevalence of cognitive disability and mobidity disability is observed in all age groups. The slope is the flattest in the young population and it gets sharper in older population. 
```{r, message=FALSE, echo=FALSE, warning=FALSE}
scat_1 
```

### Distribution of Disability by State {.tabset}
- Based on the calculated average prevalence by age, there is an increasing trend by age seen in the prevalence of any disability: 18.86% for 18-44, 29.77% for 45-64, and 44.13% for 65+. Similar increasing trends are observed in hearing disability, mobility disability, vision disability and independence disability. An obvious positive association between age and prevalence of any disability was osberved from the boxplots. 

- As shown from the barchart, the prevalence of cognitive disability is the greatest for age of 18-44 in all states in 2018. Mobidity disability is the most prevalent for age 45-64 in all states, and it's also the most prevalent for age 65+ in most states (the prevalence of hearing disability is the greatest in several states).

- As shown from the scatterplots, there is a positive association between prevalence of cognitive disability and mobidity disability, meaning that if the prelence of cognitive disability is high in one state the prevalence of mobidity disability is also likely to be high.The slopes for age 45-64 and 65+ are close whereas the curve is relatively flat for age of 18-44. 

#### For age of 18-44
```{r, message=FALSE, echo=FALSE, warning=FALSE}
map_18
```

#### For age of 45-65
```{r, message=FALSE, echo=FALSE, warning=FALSE}
map_45
```


#### For age over 65
```{r, message=FALSE, echo=FALSE, warning=FALSE}
map_65
```



